Performance for the masses: Experiments with A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-02-02 GECCO/volunteer-clouds-gecco16.tex
Revises again conclusions
Moving a part from results to there. Refs #98

% (Paloma) Mario, isn't this the future work? Should it be in conclusions? If not, just ignore this comment.
There are several validity threats to the design of this study.
Participants of the experiments were mainly members of our 
own social networks, personal colleagues, friends 
and acquaintances, therefore to extend this work other options 
should be considered in order to include 
a wider audience with different motives for participation.
Another threat is that participants were anonymous so the amount of repeated 
participations and the effects this might have had in the outcome is not
known. Another issue is that the scalability of the system was not thoroughly tested
but for the amount of participants the basic configuration 
was more than adequate, then again more experiments are needed in order
to assess the capability both with more clients and other server
configurations. Although care was taken to include representative benchmarks
of real world problems, more experiments with a broader coverage 
should allow a better understanding of the practical limitations of
the architecture and the tools chosen.

@@

In this paper two versions of a client-server architecture for volunteer and distributed
evolutionary algorithms have been evaluated. The architecture uses the browser and has been generated using {\sf
  NodIO}, based on the {\sf NodEO} evolutionary algorithm library. Volunteers are {\em in the cloud}, as stated in the title,
since they are a {\em CPU as a service} for the persons running the
experiment. In fact, in this paper we have tried to put some figures
on the real size of that {\em cloud} and how it can be used standalone
if there is no alternative, or in conjunction with other local or
cloud-based methods to add computing power in a seamless way through
the pool that {\sf NodIO} creates.

In order to establish a baseline performance, the evolutionary
algorithm was run in a desktop client-program written in JavaScript
using NodEO to solve the 40-trap function. The first experiments with
{\sf NodIO} proved that, although obtaining better performance than the
baseline was possible, it did not happen, mainly because of the
difficulty in carrying over other experiments to volunteers when the
experiment they were participating on was finished. This has been shown in the low
correlation between the number of IPs in successive experiments, and also resulted in a low number of generations allotted by users.

In the second implementation, {\sf NodIO-W$^2$}, Web Workers
were used so that several clients per browser could be executed and
they could run in the background, so that the the number of
generations per user could be improved. The number of generations per
user increased to the point that baseline performance was improved in
most cases. The architectural changes led to a high correlation
between successive experiments, so that, even as the median number of
volunteers per experiment decreased, they were probably contributing
to the experiment at the same time, so that our objective was
achieved. In general, one of the first conclusions for this paper is
that the browser technologies should be used to its full extent so
that the user time is leveraged to its full potential, and this was
achieved with {\sf NodIO-W$^2$}.

The second objective of this paper was to model the user behavior in a
first attempt to try and predict performance. As should be expected,
the model depends on the implementation, with contributions following
a GEV distribution in the case of {\sf NodIO} and a
Weibull distribution for {\sf NodIO-W$^2$}. These distributions are
close enough to each other and, in the second case, reflect the fact
that all time spent in the page is actually devoted to computing,
which is why the time spent (represented by the number of
contributions to the pool) follows a model quite similar to that found
for games or other online activities. The reverse might be true: if we
want to have returning users for the experiments, it is probable that
we should {\em gamify} the experience so that once they have done it
once, they might do it more times. In the spirit of Open Science, this
gamification might involve computing in real-time data such as the one
presented in this paper, while showing it in the same page.

In general, linking and finding correlations between user choices and
performance is an interesting avenue to explore in the future.
%  Even if
% the three previous experiments were published in a similar way, one
% obtained up to 5 times more total cycles  than the one with the least
% number of cycles. 
It is also essential to obtain volunteers as
simultaneously as possible, so it is possible that the features of the
social network in terms of real-time use will also play a big
role. Even as it is difficult to create controlled experiments in this
area, it is an interesting challenge to face in the future. % (Paloma) Wrong collocation

>>>>

@@

In this paper several experiments with a client-server architecture for volunteer and distributed
evolutionary algorithms have been performed. The architecture uses the
browser and has been generated using the {\sf
  NodIO} framework, which includes  {\sf NodEO} as an evolutionary
algorithm library. Volunteers are {\em in the cloud}, as stated in the
title, 
since they are a {\em CPU as a service} for the persons running the
experiment. In fact, in this paper we have tried to put some figures
on the real size of that {\em cloud} in case it is used by itself or
in conjunction with other local or 
cloud-based methods to add computing power in a seamless way through
the pool that {\sf NodIO} creates.

In order to establish a baseline performance, which was published in
\cite{2016arXiv160101607Manom}, the evolutionary algorithm was run in a desktop client application written in JavaScript
using NodEO to solve the 40-trap function. The first experiments in a
volunteer computing scenario proved that, although obtaining better performance than the
baseline  was possible, it did not happen, mainly because of the
difficulty in carrying over other experiments to volunteers when the
experiment they were participating on was finished. This has been shown in the low
correlation between the number of IPs in successive experiments, and
also resulted in a low number of generations allotted by users.

In the second implementation, Web Workers
were used so that several clients per browser page could be executed and
run in the background; our intention was to increase the time spent by
the users and, in fact, the number of generations per
user increased to the point that we obtained a performance better than
baseline in most cases. The architectural changes led to a high correlation
between successive experiments, so that, even as the median number of
volunteers per experiment decreased, they were probably contributing
to the experiment simultaneously, so that our objective was
achieved. In general, one of the first conclusions for this paper is
that the browser technologies should be used to its full extent so
that the user time is leveraged to its full potential, and this was
achieved using web workers.

The second objective of this paper was to model the user behavior in a
first attempt to try and predict performance. As should be expected,
the model and its parameters depends on the implementation, with contributions following
a GEV distribution with single-thread and a
Weibull distribution when web workers are used. These distributions are
close enough to each other and, in the second case, reflect the fact
that most time spent in the page is actually devoted to computing, as
opposed to presenting the chart or not doing anything,
which is why the time spent (represented by the number of
contributions to the pool) follows a model quite similar to that found
for games or other online activities. The reverse implication, from
games to volunteer computing,  might also be true: if we
want to have returning users for the experiments, it is probable that
we should {\em gamify} the experience so that once they have participated
once, they would be encouraged to do for a longer time or to share
it. In the spirit of Open Science, this 
gamification might involve computing in real-time data such as the one
presented in this paper, while showing it in the same page.

In general, linking and finding correlations between user choices and
performance is an interesting avenue to explore in the future. % (Paloma) Mario, isn't this the future work? Should it be in conclusions? If not, just ignore this comment.
% Moved to conclusions. 
There are several validity threats to the design of this study.
Participants of the experiments were mainly members of our 
own social networks, personal colleagues, friends 
and acquaintances, therefore to extend this work other options 
should be considered in order to include 
a wider audience with different motives for participation.
Another threat is that participants were anonymous so the amount of repeated 
participations and the effects this might have had in the outcome is not
known. Another issue is that the scalability of the system was not thoroughly tested
but for the amount of participants the basic configuration 
was more than adequate, then again more experiments are needed in order
to assess the capability both with more clients and other server
configurations. Although care was taken to include representative benchmarks
of real world problems, more experiments with a broader coverage 
should allow a better understanding of the practical limitations of
the architecture and the tools chosen.
%  Even if
% the three previous experiments were published in a similar way, one
% obtained up to 5 times more total cycles  than the one with the least
% number of cycles. 
It is also essential to obtain volunteers as
simultaneously as possible, so it is possible that the features of the
social network in terms of real-time use will also play a big
role. Even as it is difficult to create controlled experiments in this
area, it is an interesting challenge to face in the future. % (Paloma) Wrong collocation
